Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-06-22 src/tex/ms.tex
Finished abstract and section 1

\begin{abstract}
	We investigate the possibility of improving the accuracy of a
	phenomenological waveform model, IMRPhenomD, by jointly optimizing all the
	calibration coefficients at once, given a set of numerical relativity (NR)
	waveforms. When IMRPhenomD was first calibrated to NR waveforms, different
	parts of the waveform were calibrated separately. Using \ripple, a library
	of waveform models compatible with automatic differentiation, we can perform
	gradients-based optimization to all the waveform coefficients the same time,
	which should improve the quality of waveform by capturing correlations
	between previous separately fitted parts. We found that after recalibration,
	the median mismatch between the model and NR waveforms decrease by $50\%$.
	We further explore how different parts of the source parameter space respond
	to the optimization procedure. We find that the degree of improvement
	correlates with the spins of the source. This work shows a promising avenue
	to help understand and treat systematic error in waveform models.
\end{abstract}

@@

Many data analysis tasks in today's gravitational wave (GW) astrophysics, such
as match filtering search \citep{owen1996search, owen1999matched} and parameter 
estimation\citep{Dax:2021tsq, Islam:2022afg, zackay2018relative}, rely on accurate waveform
models. Because using numerical relativity (NR) waveforms in these tasks is
prohibitively expensive, the community has constructed waveform approximants
that can be evaluated much faster. There are three families of commonly used GW
approximants, which are effective-one-body (EOB) \citep{ossokine2020multipolar,
cotesta2020frequency, taracchini2014effective}, Numerical Relativity (NR)
surrogate \citep{islam2022surrogate, varma2019surrogate, varma2019surrogate2}
and phenomenological (Phenom) models \citep{husa2016frequency,
khan2016frequency, garcia2020multimode, pratten2021computationally}. While the
detail of construction of these models are different, they all have a set of
internal parameters that can be calibrated to NR waveforms. The quality of a
waveform model is determined by the ansatz used and the accuracy of the
calibrated parameters.

@@

The LIGO-VIRGO-KAGRA (LVK) collaboration
(\citep{LIGOScientific:2014pky,LIGOScientific:2021usb,LIGOScientific:2021djp,
VIRGO:2014yos,KAGRA:2020tym}.) has recently started their fourth observational
run on May 26, 2023. Because of the sensitivity improvement, the new run is
expected to double the size of current binary black hole (BBH) observations
\citep{abbott2020prospects}. The improved sensitivity also implies we expect to
detect individual events with a higher signal-to-noise ratio (SNR). This means
we can resolve more features in the signal, at the same time put a more
stringent requirement on the accuracy of our waveform model \citep{purrer2020gravitational, hu2022assessing}.

@@

Because of the large number of calibration parameters (often in a couple hundreds if not
more), waveform models is usually calibrated in pieces \citep{khan2016frequency, 
santamaria2010matching, pratten2021computationally}. This
ignores the correlation between different parts of the waveform model and limits
it quality. Recently, there has been efforts in rebuilding waveform models
\citep{khan2016frequency} that supports automatic differentiation (AD) 
\citep{ripple, Iacovelli:2022bbs, Iacovelli:2022mbg, Coogan:2022qxs}, which is a technique 
used to compute derivatives of functions down to machine precision
without issues of scaling up to high dimension or expression swelling. In
particular, {\ripple} \citep{ripple} exposes the calibration parameters to the
user. This allows us to make use of infrastructures that are heavily used in
machine learning, such as gradient descent and back propagation \cite{jax2018github, 
pytorch, tensorflow2015}, to improve the calibration of the waveform models.

% Put more focus on the joint optimization instead of having a set of more accurate 
% coefficients, since the method is the main point
In this paper, we investigate the possibility of further improving the accuracy
of a waveform model, IMRPhenomD \citep{khan2016frequency, husa2016frequency}, by jointly optimizing all the
calibration coefficients given a set of NR waveforms. Training on a few NR
waveforms, we demonstrate one can improve the match between IMRPhenomD and NR
waveforms over a decently sized parameter space, up to mass ratio $q=8$ We also
explore how different parts of the source parameter space (e.g. the primary and
second spins) respond to the optimization procedure, by optimizing the waveform
separately for different regions of the parameter space. This gives us hints on
whether the waveform model performs equally well in different regions of the
parameter space.

The rest of the paper is structured as follows: In Sec.~\ref{sec:method}, we
review the parameterization of the IMRPhenomD model and the mismatch function
that is used as an objective function for the calibration, followed by outlining the
specific optimization scheme used for recalibration. In
Sec.~\ref{sec:result}, we give the optimization result by comparing mismatches
of optimized waveforms with original waveforms. We also show how the
optimization result differs with waveforms of different source parameters. In
Sec.~\ref{sec:discussion}, we address the difference between our calibrating
procedure with \citep{khan2016frequency}. We also explain how reduced spin
parameterization affects the accuracy of the model.

>>>>

\begin{abstract}
	We investigate the possibility of improving the accuracy of the
	phenomenological waveform model, IMRPhenomD, by jointly optimizing all the
	calibration coefficients at once, given a set of numerical relativity (NR)
	waveforms. When IMRPhenomD was first calibrated to NR waveforms, different
	parts (i.e., the inspiral, merger, and ringdown) of the waveform were calibrated separately. 
	Using \ripple, a library of waveform models compatible with automatic differentiation, we can, for the first time,
	perform gradient-based optimization on all the waveform coefficients at the same time.
	This joint optimization process allows us to capture previously ignored correlations between separate parts of the waveform.
	We found that after recalibration, the median mismatch between the model and NR waveforms decreases by $50\%$.
	We further explore how different regions of the source parameter space respond
	to the optimization procedure. We find that the degree of improvement
	correlates with the spins of the source. This work shows a promising avenue
	to help understand and treat systematic error in waveform models.
\end{abstract}

@@

Many data analysis tasks in gravitational wave (GW) astrophysics, such
as match filtering~\citep{owen1996search, owen1999matched} and parameter 
estimation~\citep{Dax:2021tsq, Islam:2022afg, zackay2018relative, Ashton:2018jfp}, rely upon accurate waveform
models. Since the generation of numerical relativity (NR) waveforms is
prohibitively expensive, the community has constructed waveform approximants
that can be evaluated much faster. There are three families of commonly used GW
approximants: the effective-one-body (EOB) \citep{ossokine2020multipolar,
cotesta2020frequency, taracchini2014effective}, NR
surrogate \citep{islam2022surrogate, varma2019surrogate, varma2019surrogate2},
and phenomenological (Phenom) models \citep{husa2016frequency,
khan2016frequency, garcia2020multimode, pratten2021computationally}. While the
detailed construction of each model is different, they all have a set of
internal parameters that can be calibrated to NR waveforms. The quality of the
waveform model is therefore determined by the ansatz used and the accuracy of the
calibrated parameters.

@@

The LIGO-VIRGO-KAGRA (LVK) collaboration
~\citep{LIGOScientific:2014pky,LIGOScientific:2021usb,LIGOScientific:2021djp,
VIRGO:2014yos,KAGRA:2020tym} recently started their fourth observational
run on May 26, 2023. Impressively, they are
expected to double the total number of observed binary black holes (BBHs)~\citep{abbott2020prospects}.
Morover, the improved sensitivity also implies that we expect to
detect individual events with a higher signal-to-noise ratio (SNR) than ever before. This means
we can resolve more features in the signal, therefore putting more
stringent requirements on the accuracy of our waveform model \citep{purrer2020gravitational, hu2022assessing}.

@@

Because of the large number of calibration parameters (often a few hundred if not
more), waveform models are usually calibrated separately for the inspiral, merger,
and ringdown parts of the waveform \citep{khan2016frequency, 
santamaria2010matching, pratten2021computationally}. This
ignores the correlation between different parts of the waveform model and limits
it quality. Recently, there has been an effort to rebuild waveform models
\citep{khan2016frequency} using programming languages that support automatic differentiation (AD) 
\citep{ripple, Iacovelli:2022bbs, Iacovelli:2022mbg, Coogan:2022qxs}; AD is a technique 
used to compute machine precision derivatives of functions 
without the issues of scaling up to high dimension or expression swelling. In
particular, \ripple~\citep{ripple} exposes the calibration parameters to the
user. This allows us to make use of common techniques from
machine learning, such as gradient descent and back propagation \citep{jax2018github, 
pytorch, tensorflow2015}, to improve the calibration of the waveform models.

% Put more focus on the joint optimization instead of having a set of more accurate 
% coefficients, since the method is the main point
In this paper, we investigate the possibility of further improving the accuracy
of a waveform model, IMRPhenomD \citep{khan2016frequency, husa2016frequency}, by jointly optimizing all the
calibration coefficients for given a set of NR waveforms. Using a similar set of NR
waveforms to those used in \citep{khan2016frequency, husa2016frequency},
we demonstrate that one can improve the match between IMRPhenomD and NR
waveforms over a decently sized parameter space, up to a mass ratio $q=8$. We additionally
explore how different parts of the source parameter space (e.g. the primary and
second spins) respond to the optimization procedure by optimizing the waveform
separately for different regions. This can help in understanding
whether the waveform model ansatz performs equally well in different regions of the
parameter space.

The rest of the paper is structured as follows: In Sec.~\ref{sec:method}, we
review the parameterization of the IMRPhenomD model and the mismatch function
that is used as an objective/loss function for the calibration, followed by outlining the
specific optimization scheme used for recalibration. In
Sec.~\ref{sec:result}, we give the optimization result by comparing mismatches
of the optimized waveforms with the original waveforms. We also show how the
optimization result differs as a function of the source parameters. In
Sec.~\ref{sec:discussion}, we discuss the difference between our calibration
procedure and the procedure used in \citep{khan2016frequency}. We also explain how the reduced spin
parameterization affects the accuracy of the model.
